Contracts for unions and intersections are hard for Nickel

### 2021-05-11 paper.tex
Add custom contract example, small fixes

%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%

@@

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      Num,
      Bool,
      Str,
      List,
      Dyn,
      forall,
    },
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{dark-grey}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{bright-grey}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

Build systems, package managers, operating systems, cloud infrastructures, and
web services\unsure{Also continuous integration} are examples of modern complex software systems that require an
extensive and non-trivial configuration in order to make them adapted to each
different use-case. To manage the growing complexity that is then offloaded to
configurations, the declarative approach has become more and more popular,
illustrated for example by the infrastructure-as-code paradigm. This leads
configuration to play an important role in critical aspects of software
engineering, including security, availability, and maintainability.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \&
  Felleisen}.\unsure{The historical bit should probably be moved to
  related works eventually} Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts. In "first-class contracts", we include the
ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability of writing a small
validation function for, say, a URL, and having it immediately available as a
contract that can be composed with other type
constructors\unsure{Needs code to illustrate.}, is in our
opinion one of the main appeal of Nickel's contracts system. In this hybrid
system, each type constructor of the static type system - arrows, records,
foralls - must have a contract counterpart that checks at run-time that a value
is indeed a member of the corresponding type. Contracts for various extensions
of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
common extensions to the polymorphic lambda calculus, one of the most useful and
widespread turns out to be one of the most challenging: \emph{union types}.

@@

Unions are fundamental and ubiquitous building blocks of program data. A union
only holds one value whose shape is only known to be in a fixed set of
alternatives: for eample, either a boolean or a string. Wuen used in combination
with products, they can represent arbitrary tree-like structures, allowing to
model a large class of data.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data
types\unsure{This is a very misleading statement at best. Sums are
  "tagged union" sure, but that refers to an encoding a sums with
  unions, certainly not an encoding of unions with sums!} [references],
also called \emph{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

\begin{lstlisting}[language=caml]
type either = Left of int | Right of string
\end{lstlisting}

>>>>

\definecolor{OliveGreen}{RGB}{128,128,0}
%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%

@@

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      Num,
      Bool,
      Str,
      List,
      Dyn,
      forall,
    },
    keywordsprefix=\#,
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{dark-grey}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{bright-grey}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

Build systems, package managers, operating systems, cloud infrastructure,
continuous integration and web services\resolved{Also continuous integration}
are examples of modern complex software systems that require an extensive and
non-trivial configuration in order to make them adapted to each different
use-case. To manage the growing complexity that is then offloaded to
configurations, the declarative approach has become more and more popular,
illustrated for example by the infrastructure-as-code paradigm. This leads
configuration to play an important role in critical aspects of software
engineering, including security, availability, and maintainability.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
historical bit should probably be moved to related works eventually} Nickel
adopts a similar type system, with both statically typed terms, dynamically
typed terms, and first-class contracts. In "first-class contracts", we include
the ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability to write a small
validation function for, say, a port number, to have it immediately available as
a contract that can be composed with other type constructors\unsure{Needs code
to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
system.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:union-encoding}
\medskip
\small
Defines a port contract from a predicate and use it in the definition of a
\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
contract to a value, similar to `:` for static types. We assume that a
\lstinline+Host+ contract has also been defined.
\end{figure}

In this hybrid system, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that checks at run-time that
a value is indeed a member of the corresponding type. Contracts for various
extensions of the simply typed $\lambda$-calculus have been investigated
\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
Amongst all the common extensions to the polymorphic lambda calculus, one of the
most useful and widespread turns out to be one of the most challenging:
\emph{union types}.

@@

Unions are fundamental and ubiquitous building blocks of program data. A union
only holds one value whose shape is only known to be in a fixed set of
alternatives: for example, either a boolean or a string. When used in combination
with products, they can represent arbitrary tree-like structures, allowing to
model a large class of data.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data
types\unsure{This is a very misleading statement at best. Sums are
  "tagged union" sure, but that refers to an encoding a sums with
  unions, certainly not an encoding of unions with sums!} [references],
also called \emph{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and many more.

@@

\begin{lstlisting}[language=caml]
type either = Left of float | Right of string
\end{lstlisting}
